---
title: "Eberle_Covid19Final"
author: "Ben Eberle"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
```
Data Overview and Exploration

Before diving into the analysis, it’s important to understand the structure and contents of the dataset. This section will provide a summary and some basic exploratory analysis to help you get familiar with the data.

```{r import_data, include=TRUE}
# Import the data
confirmed <- read_csv("time_series_covid19_confirmed_global.csv")

# Overview of the data structure
glimpse(confirmed)

# Summary of key columns
#summary(confirmed)

# Check for missing values
missing_values <- confirmed %>%
  summarize_all(~ sum(is.na(.)))

# Print missing values summary
print(missing_values)

# Exploratory Data Analysis (EDA)
# Calculate total cases per country
country_summary <- confirmed %>%
  pivot_longer(cols = starts_with("1/"), 
               names_to = "date", 
               values_to = "confirmed_cases") %>%
  mutate(date = mdy(date)) %>%
  group_by(Country) %>%
  summarize(total_cases = sum(confirmed_cases, na.rm = TRUE),
            max_daily_cases = max(confirmed_cases, na.rm = TRUE),
            first_case_date = min(date[confirmed_cases > 0])) %>%
  arrange(desc(total_cases))

# Display the top 10 countries by total cases
head(country_summary, 10)

# Visualization: Total Cases by Country
ggplot(country_summary, aes(x = reorder(Country, -total_cases), y = total_cases)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Total Confirmed COVID-19 Cases by Country",
       x = "Country",
       y = "Total Confirmed Cases")

# View the structure of the data
#str(confirmed)

# Tidy the data
confirmed_long <- confirmed %>%
  pivot_longer(cols = starts_with("1/"), 
               names_to = "date", 
               values_to = "confirmed_cases") %>%
  mutate(date = mdy(date)) %>%
  group_by(Country, date) %>%
  summarize(confirmed_cases = sum(confirmed_cases, na.rm = TRUE)) %>%
  ungroup()

# Check the first few rows of the tidy data
head(confirmed_long)
```
Exploratory Data Analysis (EDA)

Next, we perform some exploratory data analysis to calculate total cases per country and identify which countries were most affected.
```{r analysis, include=TRUE}
# Calculate total cases per country
country_summary <- confirmed %>%
  pivot_longer(cols = starts_with("1/"), 
               names_to = "date", 
               values_to = "confirmed_cases") %>%
  mutate(date = mdy(date)) %>%
  group_by(Country) %>%
  summarize(total_cases = sum(confirmed_cases, na.rm = TRUE),
            max_daily_cases = max(confirmed_cases, na.rm = TRUE),
            first_case_date = min(date[confirmed_cases > 0])) %>%
  arrange(desc(total_cases))

# Display the top 10 countries by total cases
head(country_summary, 10)
```
Global and Country-Specific Trends

Global Trends of Confirmed COVID-19 Cases

We aggregate the data to observe the global trends in confirmed COVID-19 cases over time.
```{r global_trends, include=TRUE}
# Aggregate data to see global trends
global_trend <- confirmed_long %>%
  group_by(date) %>%
  summarize(global_confirmed_cases = sum(confirmed_cases))

# Plot global trend over time
ggplot(global_trend, aes(x = date, y = global_confirmed_cases)) +
  geom_line(color = "blue") +
  theme_minimal() +
  labs(title = "Global Trends of Confirmed COVID-19 Cases",
       x = "Date",
       y = "Total Confirmed Cases")
```
Country-Specific Case Counts

We further analyze the trends for selected countries to compare their experiences during the pandemic.
```{r country_counts, include=TRUE}
# Select a few key countries for comparison
selected_countries <- c("US", "Italy", "South Korea", "Germany", "Brazil")

country_trend <- confirmed_long %>%
  filter(Country %in% selected_countries)

# Plot country-specific trends
ggplot(country_trend, aes(x = date, y = confirmed_cases, color = Country)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Country-Specific COVID-19 Case Counts",
       x = "Date",
       y = "Confirmed Cases",
       color = "Country")
```
Modeling: Impact of Lockdowns on Growth Rate of Cases

Calculating Daily Growth Rate

To understand the impact of lockdowns, we calculate the daily growth rate of confirmed cases for each country.
```{r daily_growth, include=TRUE}
# Calculate daily growth rate
confirmed_long <- confirmed %>%
  pivot_longer(cols = starts_with("1/"), 
               names_to = "date", 
               values_to = "confirmed_cases") %>%
  mutate(date = mdy(date)) %>%
  group_by(Country) %>%
  arrange(date) %>%
  mutate(daily_growth_rate = (confirmed_cases - lag(confirmed_cases)) / lag(confirmed_cases) * 100) %>%
  ungroup()

# Remove rows with NA, NaN, or Inf in daily_growth_rate
confirmed_long <- confirmed_long %>%
  filter(is.finite(daily_growth_rate))
```

We model the impact of lockdowns on the daily growth rate using a linear model, focusing on a selected group of countries.
```{r lockdown_impact_f, include=TRUE}
# Calculate daily growth rate
confirmed_long <- confirmed_long %>%
  group_by(Country) %>%
  arrange(date) %>%
  mutate(daily_growth_rate = (confirmed_cases - lag(confirmed_cases)) / lag(confirmed_cases) * 100) %>%
  ungroup()

# Remove rows with NA, NaN, or Inf in daily_growth_rate
confirmed_long <- confirmed_long %>%
  filter(is.finite(daily_growth_rate))

# Check for any remaining problematic values
sum(is.na(confirmed_long$daily_growth_rate)) # should be 0

# Example lockdown dates for these countries
lockdown_dates <- tibble(
  Country = c("US", "Italy", "South Korea", "Germany", "Brazil"),
  lockdown_date = as.Date(c("2020-03-19", "2020-03-09", "2020-02-23", "2020-03-22", "2020-03-24"))
)

# Select the countries that were included in the model training
selected_countries <- c("US", "Italy", "South Korea", "Germany", "Brazil")

# Filter the dataset to only include these countries
model_data <- confirmed_long %>%
  filter(Country %in% selected_countries)

# Merge with lockdown dates
model_data <- model_data %>%
  left_join(lockdown_dates, by = "Country") %>%
  mutate(post_lockdown = ifelse(date >= lockdown_date, 1, 0))

# Fit the model again with the filtered data
model <- lm(daily_growth_rate ~ post_lockdown + Country, data = model_data)
summary(model)

# Predicting with the same countries
model_data <- model_data %>%
  mutate(predicted_growth_rate = predict(model, model_data))

# Visualization
ggplot(model_data, aes(x = date, y = daily_growth_rate, color = Country)) +
  geom_line() +
  geom_point(aes(y = predicted_growth_rate), shape = 1, size = 3) +
  theme_minimal() +
  labs(title = "Effectiveness of Lockdowns on COVID-19 Growth Rate",
       x = "Date",
       y = "Daily Growth Rate (%)",
       color = "Country")
```
Bias Analysis

4.1 Data Coverage and Selection Bias

Geographical Bias:

	•	Description: The analysis focuses on a subset of countries, including the US, Italy, South Korea, Germany, and Brazil. While these countries were significantly impacted by COVID-19, they represent only a fraction of the global landscape.
	•	Implications: This selection may lead to an overestimation or underestimation of global trends. For instance, countries with different healthcare infrastructures, government responses, and public compliance levels might show different patterns in case growth and response to lockdowns. By focusing on these selected countries, we may miss out on understanding the full spectrum of pandemic responses and outcomes across the world.

Temporal Bias:

	•	Description: The dataset covers confirmed COVID-19 cases from January 22, 2020, to March 9, 2023. This period includes multiple waves of the pandemic, various public health interventions, and the introduction of vaccines.
	•	Implications: The timing of lockdowns and other interventions varied widely across countries. Early interventions might show different impacts compared to those implemented later in the pandemic. Additionally, as the pandemic evolved, factors such as changes in testing capacity, public behavior, and mutations of the virus could influence the effectiveness of interventions. Analyzing the entire period without accounting for these evolving factors may lead to oversimplified conclusions.

Selection of Variables:

	•	Description: The model uses only two variables: whether the date is post-lockdown and the country itself. This approach assumes that lockdowns and country differences are the primary determinants of changes in the daily growth rate.
	•	Implications: This may neglect other crucial factors such as public compliance, healthcare capacity, economic conditions, social behavior, and subsequent public health measures (e.g., mask mandates, vaccination campaigns). The model’s simplicity might mask the complexity of factors that contribute to the spread of COVID-19, leading to biased or incomplete conclusions.

4.2 Modeling Bias

Linear Assumptions:

	•	Description: The linear model assumes a straightforward, proportional relationship between the implementation of lockdowns and the reduction in daily growth rates. It also assumes that the impact of lockdowns is uniform across all countries.
	•	Implications: In reality, the relationship between interventions like lockdowns and case growth may be non-linear. Factors such as public compliance, healthcare system capacity, and the timing of interventions can create non-linear effects. For instance, early and strict lockdowns might have a more significant impact compared to delayed or less stringent measures. The linear assumption might lead to an underestimation of the effectiveness of early, stringent lockdowns or an overestimation of the impact of late or lax lockdowns.

Lockdown Effectiveness:

	•	Description: The model attributes changes in the daily growth rate primarily to the presence of a lockdown. Other concurrent interventions (e.g., increased testing, contact tracing, public awareness campaigns) and external factors (e.g., public events, travel restrictions) are not considered in the model.
	•	Implications: The effectiveness of lockdowns could be confounded by these other factors. For example, a reduction in growth rate might coincide with a lockdown but could be driven by other measures such as mass testing or natural herd immunity. By not accounting for these factors, the model may overstate the effectiveness of lockdowns alone.

4.3 Data Quality and Reporting Bias

Reporting Bias:

	•	Description: The dataset relies on reported confirmed cases, which can be influenced by the testing capacity, reporting practices, and transparency of each country. Some countries might have underreported cases due to limited testing, political reasons, or delays in data collection.
	•	Implications: If certain countries underreport cases, the analysis might show lower growth rates or smaller outbreaks than actually occurred. This could skew the model’s results, leading to biased conclusions about the effectiveness of interventions in those countries. Additionally, countries with better testing and reporting infrastructure might appear to have higher cases, not necessarily because the outbreak was worse, but because it was better documented.

Data Completeness:

	•	Description: Missing or incomplete data, especially in the early days of the pandemic, could lead to inaccurate growth rate calculations. For example, if case counts are not reported on certain days, this could artificially inflate or deflate the calculated growth rates.
	•	Implications: Incomplete data can introduce significant errors into the model, especially if these gaps occur during critical periods (e.g., just before or after a lockdown). This could lead to incorrect inferences about the timing and effectiveness of interventions.

4.4 Analytical Bias

Simplification of Complex Relationships:

	•	Description: The analysis simplifies the complex dynamics of the COVID-19 pandemic by focusing primarily on the impact of lockdowns on case growth rates.
	•	Implications: The pandemic’s progression was influenced by a myriad of factors, including social behaviors, economic conditions, public health infrastructure, and government policies beyond lockdowns. Simplifying the analysis to focus on lockdowns alone may overlook the interaction between these factors, leading to an incomplete